Cosmological Evidences of Dark Matter through the CMB

### 2019-02-19 paper/main.tex
phenomenology of the cmb

@@

\section{Introduction}
Many astronomical and cosmological observations suggest that there is a missing form of matter that interacts gravitationally. 
In order to explain such observations, scientists have proposed the existence of Dark Matter, a new form of matter that does not interact electromagnetically but gravitationally.
However, the fundamental nature of Dark Matter is not well understood yet, and whether Dark Matter is truly a new form of matter or the effect of new laws of gravity is still matter of debate.\\
One of the most compelling evidence for the existence of Dark Matter comes from the analysis of the Cosmic Microwave Background (CMB).
In this paper we explain how the  analysis of the CMB is linked to the energy-matter content of our Universe, and, within this framework, we study the role of Dark Matter.
The aim of this paper is to give a short and concise overview of the cosmological evidence of Dark Matter given by the observations of the CMB.
We are interested in providing few but fundamental ingredients to understand the importance of the CMB for Dark Matter to a reader that is not familiar with the topic.
\\

@@

The Cosmic Microwave Background is the electromagnetic radiation coming from the hot plasma of the early stages of the Universe.
When the Universe, made of hot plasma, reached a temperature low enough to form stable hydrogen $e^- + p^+  \rightarrow H + \gamma$ the number density of the free electrons dropped, causing the Thomson scattering $e^- + \gamma  \rightarrow e^- + \gamma$ to be inefficient and the photons to decouple. They have since streamed freely through the universe and are today observed as the Cosmic Microwave Background \citep{LecturesPdf}.\\

-practical Motivation for studyinbg the cmb: nice spectrum black body spectrum: physics well know, very intense radiation\\
which info can we get from the cmb: Big bang, maater and energy content of the universe---dark matter is a big part of the matter content it influenced the spectrum  of the CMB ---- so it played a role.\\
-\textbf{structure of the paper} \\
--The experimental results from the observations of the Cosmic Microwave Background provide important details about energy-matter content of the Universe.

@@

\dd s ^2 = -\dd t^2 + a^2 (t) \qty( \frac{1}{1-k \, r^2} \dd r^2 + r^2 \dd \Omega ^2 )  

@@

where $k$ is the spatial curvature and $a(t)$ is the scale factor (we set the light speed $c=1$).

@@

R_{\mu \nu} - \dfrac{1}{2} R \, g _{\mu \nu} + \Lambda g_{\mu \nu}= 8 \pi G \, T_{\mu \nu}

@@

T^{\mu \nu } = (p + \rho)u ^{\mu} u^{\nu}+p g^{\mu \nu} 

@@

H^2 = \qty(\frac{\dot{a}}{a})^2 = \dfrac{8 \pi G}{3} \rho -\dfrac{k}{a^2} + \frac{\Lambda}{3}

@@

\dv[]{}{t}\qty(\rho a^3) + p \dv[]{}{t} \qty(a^3)  = 0

@@

@@

\dfrac{-k}{H_0 ^2} a^{-2} + \dfrac{\Lambda}{3 H_0 ^2}
]

@@

\Omega_\Lambda = \dfrac{\Lambda}{3} \qquad \Omega_k = -  \dfrac{k}{H_0 ^2}
\\

@@

\section{CMB theo analysis}

\begin{itemize}
\item hydrodynamics \citep{huLectureNotesCMB2008}
page 14 eq 49 to page 17 eq 71\\
skip doppler effect
\item gravito acoustic oscillations page 19 up to eq 82 , justify briefly constant potential in page 20
% if stress perturbations are negligible compared with density perturbations ( δp ≪ δρ ) then the potential will remain constant in periods where the background equation of state p/ρ is constant
\item page 20 and 21 up to eq92 important comment 
%The consequence is that overdense regions where Ψ is negative (potential wells) are cold spots in the effective temperature.
\item baryonic effect
\end{itemize}

>>>>

% for comments
\usepackage{verbatim}

@@

\section{Introduction}
Many astronomical and cosmological observations suggest that there is a missing form of matter that interacts gravitationally with baryonic matter. 
In order to explain such observations, scientists have proposed the existence of Dark Matter, a new form of matter that does not interact electromagnetically but gravitationally.
The fundamental nature of Dark Matter is not well understood yet, and whether Dark Matter is truly a new form of matter or the effect of new laws of gravity is still matter of debate.\\
One of the most compelling evidence for the existence of Dark Matter comes from the analysis of the Cosmic Microwave Background (CMB).
In this paper we explain how the analysis of the CMB is linked to the energy-matter content of our Universe, and we study the role of Dark Matter on the power spectrum of the CMB within the $\Lambda$CDM model.
The aim of this paper is to give a short and concise overview of the cosmological evidence of Dark Matter given by the observations of the CMB.
We are interested in providing few but fundamental ingredients to understand the importance of the CMB for Dark Matter to a reader that is not familiar with the topic.
\\

@@

The Cosmic Microwave Background is the electromagnetic radiation coming from the hot plasma of the early stages of the Universe.
As the Universe was expanding, it cooled down to a temperature low enough to form stable hydrogen $e^- + p^+  \rightarrow H + \gamma$. The number density of the free electrons dropped with the expansion causing the Thomson scattering $e^- + \gamma  \rightarrow e^- + \gamma$ to be inefficient and the photons to decouple. They have since streamed freely through the universe and are today observed as the Cosmic Microwave Background \citep{LecturesPdf}.\\
The Cosmic Microwave Background is ideal for studying the content of our Universe, because it is the best measured Planck spectrum that we know.\\
Since we know very well the 
-practical Motivation for studyinbg the cmb: nice spectrum black body spectrum: physics well know, very intense radiation\\
which info can we get from the cmb: Big bang, maater and energy content of the universe---dark matter is a big part of the matter content it influenced the spectrum  of the CMB ---- so it played a role.\\
-\textbf{structure of the paper} \\
--The experimental results from the observations of the Cosmic Microwave Background provide important details about energy-matter content of the Universe.\\
\paragraph{sctructure}
We first give an historical overview of the discovery and measurements of the CMB, then we discuss the main assumptions of the $\Lambda$CDM model and the link between the content of our universe and the cosmological model. Afterwards, the phenomenology of the hot plasma is explained with a simplified treatment in order to understand the power spectrum of the CMB.
We conclude with the discussion of the cmb power specctrum and the role of dark matter.

@@

\dd s ^2 = -c ^2 \dd t^2 + a^2 (t) \qty( \frac{1}{1-k \, r^2} \dd r^2 + r^2 \dd \Omega ^2 )  

@@

where $k$ is the spatial curvature and $a(t)$ is the scale factor.

@@

R_{\mu \nu} - \dfrac{1}{2} R \, g _{\mu \nu} + \Lambda g_{\mu \nu}= \dfrac{8 \pi G}{c^4} \, T_{\mu \nu}

@@

T^{\mu \nu } = (p + \rho c^2)u ^{\mu} u^{\nu}+p g^{\mu \nu} 

@@

H^2 = \qty(\frac{\dot{a}}{a})^2 = \dfrac{8 \pi G}{3} \rho -\dfrac{k \, c^2}{a^2} + \frac{\Lambda \, c^2}{3}

@@

\dv[]{}{t}\qty(\rho c^2 a^3) + p \dv[]{}{t} \qty(a^3)  = 0

@@

Solving for radiation $\rho c^2 = p/3$
\[
\dot{\rho}  + 4 \frac{ \dot{a}}{a} \rho =0
\]
 And solving for matter $\rho c^2 << p$
 \[
 \dot{\rho}  + 3 \frac{ \dot{a}}{a} \rho =0
 \]

@@

\dfrac{-k \, c^2}{H_0 ^2} a^{-2} + \dfrac{\Lambda \, c^2}{3 H_0 ^2}
] \notag

@@

\Omega_\Lambda = \dfrac{\Lambda c^2}{3} \qquad \Omega_k = -  \dfrac{k c^2}{H_0 ^2} \qquad

@@

\section{Phenomenology of the Fluctuations}
The Cosmic Microwave Background is the image of the fluctuations of a dense and hot fluid made of matter and radiation.
Before the release of the CMB at $z > 1100$ the electromagnetic radiation is firmly coupled (both energetically and kinetically) to the baryons and it is possible to describe the behavior of such a system by using solving the equations of relativistic hydrodynamics and the Einstein's field equations (\ref{einstein_eq}) with small density $\delta \rho$, pressure $\delta p$ and gravitational potential perturbations $\delta \Phi$.
The full general-relativistic description of the photon-matter fluid requires a complicated treatment[BAUMANN].
However, it is possible to qualitatively discuss the phenomena occurring in the CMB fluid and the effects on the temperature fluctuations $\Theta = \delta T / T_0$.
\subsection{Sachs-Wolfe effect}
As a consequence of the Einstein's equivalence principle, photons that escape from a gravitational field are redshifted.
If we consider the fluctuations of the CMB on large scales, the photons that tries to escape from the gravitational well created by matter are redshifted.

\subsection{Acoustic oscillations}
At the time of recombination
Dark Matter is randomly distributed
if a portion of photon-baryon fluid is in a potential well of the dark matter, it will fall to the center of the well
gravity compresses the photon-baryon fluid
pressure starts to rise
the pressure is sufficient to cause the fluid to expand outward
the pressure drops until gravity causes the photon-baryon fluid to fall inward again

\subsection{Damping}
The temperature fluctuations are damped due to collisions between the baryons and the photons that dissipate energy, and the expansion of the universe.

\begin{comment}
    This text will not be displayed.

 by starting from the continuity and Euler equation for a photon fluid with  number density $n$, energy density $u$ and pressure $p$:
\begin{equation}
\label{hydro}
\dot{n} + n_0 \vec{\nabla} \cdot \vec{v} = 0 \qquad
\dot{\vec{v}} = - c^2 \dfrac{\vec{\nabla} \delta p}{p_0 + u_0} - \vec{\nabla} \delta \Phi
\end{equation}
where $v$ is the streaming velocity of the perturbations and the subscript $_0$ indicates the background average values of th correspondent quantities.\\
For a photon gas we know that
\begin{equation}
n \propto T ^3 \qquad u \propto T^4 \qquad p  = \dfrac{u}{3}
\end{equation}
So, the temperature fluctuations are related to the number density, energy density and pressure as follow:
\begin{equation}
\dfrac{\delta n}{n_0} = 3 \Theta \qquad
\dfrac{\delta u}{u_0} = 4 \Theta = \dfrac{\delta p}{p_0}
\end{equation}
We can rewrite eq.(\ref{hydro}) in terms of the temperature fluctuations:
\begin{equation}
\dot{\Theta} + \nabla \cdot \vec{v} = 0 \qquad \dot{\vec{v}} = - c^2 \vec{\nabla} \Theta - \vec{\nabla} \delta \Phi
\end{equation}
where we used $p_0 + u_0 = 4 p_0$

\begin{itemize}
\item hydrodynamics \citep{huLectureNotesCMB2008}
page 14 eq 49 to page 17 eq 71\\
skip doppler effect
\item gravito acoustic oscillations page 19 up to eq 82 , justify briefly constant potential in page 20
% if stress perturbations are negligible compared with density perturbations ( δp ≪ δρ ) then the potential will remain constant in periods where the background equation of state p/ρ is constant
\item page 20 and 21 up to eq92 important comment 
%The consequence is that overdense regions where Ψ is negative (potential wells) are cold spots in the effective temperature.
\item baryonic effect
\end{itemize}
\end{comment}
